---
title: 'Unit 3 Lecture 3: Ridge regression'
date: "October 12, 2021"
output: pdf_document
---

```{r, echo = FALSE, message = FALSE}
library(tidyverse)
```

**NOTE: This R demo has been updated since it was presented in class on October 12. Through the use of the `glmnetUtils` package, it is no longer necessary to separately construct `X` and `Y` to pass into `cv.glmnet`, or to scale the `X` matrix manually, or to remove the intercept manually. You can use `cv.glmnet` in much the same way you have been using `lm` and `glm`: by specifying a formula and supplying a data frame.**

In this R demo, we will learn about the `glmnet` and `glmnetUtils` packages and how to run a cross-validated ridge regression using the `cv.glmnet()` function. 

First, let's install the `glmnet` and `glmnetUtils` packages:
```{r}
# install.packages("glmnet")
# install.packages("glmnetUtils")
```
Next, we load the `glmnetUtils` package:
```{r}
library(glmnetUtils)
```

Let's also source a function called `plot_glmnet` to help us plot our results:
```{r}
source("../../functions/plot_glmnet.R")
```

We will be applying ridge regression to study the effect of 97 socioeconomic factors on violent crimes per capita based on data from 90 communities in Florida:
```{r, message = FALSE}
crime_data = read_csv("../../data/CrimeData_FL.csv")
crime_data
```

Let's split the data into training and testing, as usual:
```{r}
set.seed(471)
train_samples = sample(1:nrow(crime_data), 0.8*nrow(crime_data))
crime_data_train = crime_data %>% filter(row_number() %in% train_samples)
crime_data_test = crime_data %>% filter(!(row_number() %in% train_samples))
```

# Running a cross-validated ridge regression

We call `cv.glmnet` on `crime_data_train`:
```{r}
ridge_fit = cv.glmnet(violentcrimes.perpop ~ .,  # formula notation, as usual
                      alpha = 0,                 # alpha = 0 for ridge
                      nfolds = 10,               # number of folds
                      data = crime_data_train)   # data to run ridge on
```

A few things to note:

- the sequence of penalty parameters is automatically chosen for you
- `alpha = 0` means "ridge regression" (we'll discuss other values of alpha next lecture)
- `nfolds` specifies the number of folds for cross-validation
- the columns of the matrix `X` are being standardized for you behind the scenes;
there is no need to standardize yourself

# Inspecting the results

The `glmnet` package has a very nice `plot` function to produce the CV plot:
```{r}
plot(ridge_fit)
```

The `ridge_fit` object has several fields with information about the fit:
```{r}
# lambda sequence
head(ridge_fit$lambda)
# CV estimates
head(ridge_fit$cvm)
# CV standard errors
head(ridge_fit$cvsd)
# lambda achieving minimum CV error
ridge_fit$lambda.min
# lambda based on one-standard-error rule
ridge_fit$lambda.1se
```

To get the fitted coefficients at the selected value of lambda:
```{r}
coef(ridge_fit, s = "lambda.1se") %>% head()
coef(ridge_fit, s = "lambda.min") %>% head()
```

To visualize the fitted coefficients as a function of lambda, we can make a plot of the coefficients like we saw in class. To do this, we can use the `plot_glmnet` function, which by default shows a dashed line at the lambda value chosen using the one-standard-error rule:
```{r}
plot_glmnet(ridge_fit, crime_data_train) # NOTE: MUST PASS IN THE DATA AS WELL
                                         #  AS THE FIT OBJECT
```

If we want to annotate the features with the top few coefficients, we can use the `features_to_plot` argument:

```{r}
plot_glmnet(ridge_fit, crime_data_train, features_to_plot = 7)
```

To interpret these coefficient estimates, recall that they are for the *standardized* features.

# Making predictions

To make predictions on the test data, we can use the `predict` function (which we've seen before):
```{r}
ridge_predictions = predict(ridge_fit, 
                            newdata = crime_data_test,
                            s = "lambda.1se") %>% as.numeric()
ridge_predictions
```
We can evaluate the root-mean-squared-error as before:
```{r}
RMSE = sqrt(mean((ridge_predictions - crime_data_test$violentcrimes.perpop)^2))
RMSE
```

# Ridge logistic regression

We can also run a ridge-penalized logistic regression. Let's try it out on `default_data`.
```{r}
# load data, convert default to binary
default_data = ISLR2::Default %>% 
  as_tibble() %>% 
  mutate(default = as.numeric(default == "Yes"))
# split into train and test
set.seed(471)
train_samples = sample(1:nrow(default_data), 0.8*nrow(default_data))
default_train = default_data %>% filter(row_number() %in% train_samples)
default_test = default_data %>% filter(!(row_number() %in% train_samples))
```

To run the logistic ridge regression, we call `cv.glmnet` as before, adding the argument `family = binomial` to specify that we want to do a logistic regression and the argument `type.measure = "class`: to specify that we want to use the misclassification error during cross-validation.
```{r}
ridge_fit = cv.glmnet(default ~ .,             # formula notation, as usual
                      alpha = 0,               # alpha = 0 means ridge
                      nfolds = 10,             # number of CV folds
                      family = "binomial",     # to specify logistic regression
                      type.measure = "class",  # use misclassification error in CV
                      data = default_train)    # train on default_train data
```

We can then take a look at the CV plot and the trace plot as before:
```{r}
plot(ridge_fit)
plot_glmnet(ridge_fit, default_train, features_to_plot = 4)
```

To predict using the fitted model, we can use the `predict` function again, this time specifying `type = "response"` to get the predictions on the probability scale (as opposed to the log-odds scale).
```{r}
probabilities = predict(ridge_fit,               # fit object
                        newdata = default_test,  # new data to test on
                        s = "lambda.1se",        # which value of lambda to use
                        type = "response") %>%   # to output probabilities
  as.numeric()                                   # convert to vector
head(probabilities)
```

We can threshold the probabilities to get binary predictions as we did with regular logistic regression.